Solving the maximum independant set problem by generic programming einsum networks

### 2021-06-15 paper/paper.tex
new data and new introduction

\section{Introduction}
Exact algorithms that solves the MIS problem in $1.1893^n n ^{O(1)}$ time ~\cite{Xiao2017}.

@@

\section{Automated branching}
% check http://www.tcs.rwth-aachen.de/independentset/ for more rules
Branching rules can be automatically discovered by contracting the tropical einsum network for a subgraph $R \subseteq G$.
Let us denote the resulting tropical tensor of rank $|C|$ as $A$, where $C$ is the set of boundary vertices defined as $C := \{c | c\in R \land c \in G\backslash R\}$ and $|C|$ the size of $C$.
Each tensor entry $A_{\sigma}$ is a local maximum independant set size with a fixed boundary configuration $\sigma \in \{0,1\}^{|C|}$ by marginalizing the inner degrees of freedom.
If we are only interested in finding a single maximum independent set rather than enumerating all possible solutions,
this tensor can be further ``compresed'' by setting some entries to tropical zero.
Let us define a relation of \textit{less restrictive} as
\begin{align}
(\sigma_a \prec \sigma_b) := (\sigma_a \neq \sigma_b) \land (\sigma_a \leq^\circ \sigma_b)
\end{align}
where $\leq^\circ$ is the Hadamard less or equal operations.
We say an entry $A_{\sigma_a}$ is ``better'' than $A_{\sigma_b}$ if
\begin{align}
(\sigma_a \prec \sigma_b) \land (A_{\sigma_a} \geq A_{\sigma_b}),\label{eq:compress}.
\end{align}
If we remove such $A_{\sigma_b}$, the contraction over the whole graph is guaranted to give the same maximum independant set size.
It can be seen by considering two entries with the same local maximum independent set sizes and different boundary configurations as shown in \Fig{fig:compressrule} (a) and (b).
If we have $\sigma_b \cup \overline{\sigma_b}$ being one of the solutions for maximum independant sets in $G$, then $\sigma_a \cup \overline{\sigma_b}$ is another solution giving the same $\alpha(G)$.
Hence, we can set $A_{\sigma_b}$ to tropical zero safely.
We name this entry elimination principle as the \textit{least restrictive principle}, that is,
among boundary configurations with equal local maximum independent sizes,
we only retain those least restrictive (less ones at the boundary) to exterial configurations.

@@

This rule states that if $v$ is not in $M$, there exists an MIS $I$ that $M(v)\notin I$.
otherwise, there must be one of $N(v)$ in the MIS (\textit{local maximum rule}).
If $w$ is in $I$, then none of $N(v) \cap N(w)$ is in $I$, then there must be one of node in the clique $N(v)\backslash N(w)$ in $I$ (\textit{local maximum rule}),
since clique has at most one node in the MIS, the tensor compression will eliminate this solution by moving the occuppied node to the interior.
Hence, the \textit{least restrictive principle} captures the mirror rule.

>>>>

\section{Introduction}
Exact algorithms that solves the MIS problem in $1.1893^n n ^{O(1)}$ time ~\cite{Xiao2017}.
Previous dynamic programming approach~\cite{Fomin2013} can reduce the complexity of computing to $2^{tw(G)}$.

\section{A short introduction to einsum networks}

The word ``einsum'' is a shorthand for Einstein's summation, but it is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of multiplication between two matrices to the contraction between multiple tensors.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$,
or $C_i^k = A_i^j B_j^k$ in the Einstein's original notation, where the paired subscript and superscript $j$ is a dummy index summed over.
An example of tensor networks is $C_i^l = A_{ij}^k B^l_k V^j$.
One can map a tensor network to a muti-graph with open edges by viewing a tensor in the expression on the right hand side as a vertex in a graph, a label pairing two tensors as an edge, and the remaining labels as open edges.
This is because a label in a tensor network representation always appear exactly twice.

\centerline{\begin{tikzpicture}
    \def\dx{0};
    \def\r{0.25}
    \def\ax{0}
    \def\ay{0}
    \def\bx{1}
    \def\by{1}
    \def\cx{1}
    \def\cy{-1}
    \filldraw[fill=black] (\ax+\dx,\ax) circle [radius=\r];
    \filldraw[fill=black] (\bx+\dx,\by) circle [radius=\r];
    \filldraw[fill=black] (\cx+\dx,\cy) circle [radius=\r];
    \draw [black,thick] (\ax+\dx,\ay) .. controls (0.5*\ax+0.5*\bx+\dx+0.2,0.5*\ay+0.5*\by-0.2) .. (\bx+\dx,\by);
    \draw [black,thick] (\ax+\dx,\ay) .. controls (0.5*\ax+0.5*\bx+\dx-0.2,0.5*\ay+0.5*\by+0.2) .. (\bx+\dx,\by);
    \draw [black,thick] (\ax+\dx,\ay) -- (\cx+\dx,\cy);
    \draw [black,thick] (\ax+\dx,\ay) -- (\ax+\dx-0.6,\ay);
    \draw [black,thick] (\bx+\dx,\by) -- (\bx+\dx+0.6,\by);
    \node[color=white] at (\ax+\dx,\ax) {A};
    \node[color=white] at (\bx+\dx,\by) {B};
    \node[color=white] at (\cx+\dx,\cy) {V};
\end{tikzpicture}}

Numpy programmers make a generalization to this notation by not restricting the number of times a label is used by tensors.
For example, $C_{ijk} = A_{jkm} B_{mia} V_{jm}$ is a elgal einsum representation, all indices not appearing in the output are summed over, i.e. it represents $C_{ijki} = \sum_{ma}A_{jkm} B_{mia} V_{jm}$.
Whether the index appear as a superscript or a subscript makes no sense now.
The graphical representation of an einsum is called a hypergraph, where an edge can be shared by any number of nodes.

\centerline{\begin{tikzpicture}
    \def\dx{0};
    \def\r{0.25}
    \def\ax{0}
    \def\ay{0}
    \def\bx{1}
    \def\by{1}
    \def\cx{1}
    \def\cy{-1}
    \filldraw[fill=black] (\ax+\dx,\ax) circle [radius=\r];
    \filldraw[fill=black] (\bx+\dx,\by) circle [radius=\r];
    \filldraw[fill=black] (\cx+\dx,\cy) circle [radius=\r];
    \draw[color=blue,thick] (\bx+\dx,\by) circle [radius=\r+0.1];
    \node[color=blue] at (\bx+\dx-0.6,\bx) {i};
    \draw[color=cyan,dashed,thick] (\bx+\dx,\by) circle [radius=\r+0.25];
    \node[color=cyan] at (\bx+\dx+0.5,\bx-0.5) {a};
    \draw[color=violet,thick,dashed] (\ax+\dx,\ay) circle [radius=\r+0.1];
    \node[color=violet] at (\ax+\dx+0.4,-0.4) {k};
    \draw[color=black,thick] (0.5+\dx,0) circle [radius=1.7];
    \node[color=black] at (0.5+\dx-1.9,0) {m};
    \draw[color=red,thick,rotate=135,dashed] (-0.6,0) ellipse (1.2 and 0.5);
    \node[color=red] at (\ax+\dx,-1.0) {j};
    \node[color=white] at (\ax+\dx,\ax) {A};
    \node[color=white] at (\bx+\dx,\by) {B};
    \node[color=white] at (\cx+\dx,\cy) {V};
\end{tikzpicture}}

In the main text, we stick to the einsum notation, although to most physicists, they are more familiar with tensor networks,
and an einsum network can be translated to an equivalent tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
We do not use the language of tensor network because it can sometime increase the contraction complexity of a graph. We will show an example in the appendix.

@@

\section{Automated branching}
% check http://www.tcs.rwth-aachen.de/independentset/ for more rules
Branching rules can be automatically discovered by contracting the tropical einsum network for a subgraph $R \subseteq G$.
Let us denote the resulting tropical tensor of rank $|C|$ as $A$, where $C$ is the set of boundary vertices defined as $C := \{c | c\in R \land c \in G\backslash R\}$ and $|C|$ the size of $C$.
Each tensor entry $A_{\sigma}$ is a local maximum independant set size with a fixed boundary configuration $\sigma \in \{0,1\}^{|C|}$ by marginalizing the inner degrees of freedom.
If we are only interested in finding a single maximum independent set rather than enumerating all possible solutions,
this tensor can be further ``compresed'' by setting some entries to tropical zero.
Let us define a relation of \textit{less restrictive} as
\begin{align}
(\sigma_a \prec \sigma_b) := (\sigma_a \neq \sigma_b) \land (\sigma_a \leq^\circ \sigma_b)
\end{align}
where $\leq^\circ$ is the Hadamard less or equal operations.
We say an entry $A_{\sigma_a}$ is ``better'' than $A_{\sigma_b}$ if
\begin{align}
(\sigma_a \prec \sigma_b) \land (A_{\sigma_a} \geq A_{\sigma_b}),\label{eq:compress}.
\end{align}
If we remove such $A_{\sigma_b}$, the contraction over the whole graph is guaranted to give the same maximum independant set size.
It can be seen by considering two entries with the same local maximum independent set sizes and different boundary configurations as shown in \Fig{fig:compressrule} (a) and (b).
If we have $\sigma_b \cup \overline{\sigma_b}$ being one of the solutions for maximum independant sets in $G$, then $\sigma_a \cup \overline{\sigma_b}$ is another solution giving the same $\alpha(G)$.
Hence, we can set $A_{\sigma_b}$ to tropical zero safely.
%among boundary configurations with equal local maximum independent sizes,
%we only retain those least restrictive (less ones at the boundary) to exterial configurations.

@@

This rule states that if $v$ is not in $M$, there exists an MIS $I$ that $M(v)\notin I$.
otherwise, there must be one of $N(v)$ in the MIS (\textit{local maximum rule}).
If $w$ is in $I$, then none of $N(v) \cap N(w)$ is in $I$, then there must be one of node in the clique $N(v)\backslash N(w)$ in $I$ (\textit{local maximum rule}),
since clique has at most one node in the MIS, by moving the occuppied node to the interior, we obtain a ``better'' solution.
%Hence, the \textit{least restrictive principle} captures the mirror rule.
